### 2024-08-07 manuscript_ten-simple-rules-writing-retreat.tex
finalised the first draft

Writing is an integral part of academic work. No matter in your career stage, your scholarly output will advance scientific knowledge and count in your career development, getting grants or landing a job. However, in multi-tasking admin, teaching, fieldwork, among other activities, we frequently put aside how important it is to take the proper time to wrap it all up in a written document (e.g. thesis, scientific paper). Scientific writing has its technical tips and tricks, but ultimately, it is a creative act. Yet, the lack of time to think critically, create knowledge and transfer it into words is a significant complaint in academia [~\cite{menzies}]. Therefore, we have to actively disengage from other tasks to enable us to properly engage with the act of writing [~\cite{murray2013}]. In this context, writing retreats have been shown as effective opportunities to unplug from daily work routines, allowing scientists to excel in their passion: producing science.

@@

Here, we build on these experiences of student-led writing retreats and lay out Ten Simple Rules for organising an effective retreat. Althuogh the authors are/were all PhD candidates at the time of the retreats, the rules herein should be helpful for research groups and academics more broadly. We cover steps to facilitate the planning/execution of pre- (Rules 1--4), during (Rules 5--9), and pos- (Rule 10) writing retreat actions, but not writing techniques per se. There are many resources on academic writing that we will not delve into but encourage their reading (e.g. [~\cite{turbek2016, hotaling2020, carter2020, cargill2021}]), including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}]. Otherwise, you may consider writing a boring contribution instead [~\cite{sand2007}].

@@

The schedule for each day was roughly between 9 am and 6 pm and was provided in advance of the retreat so participants knew what to expect for dedicated work time (see \nameref{S1_Appendix}). A different leader was designated for each day to lead discussions and keep the time for writing blocks (between 60 and 120 min) and breaks (between 30 and 60 min). We started each morning together by writing for 15 minutes using prompts that varied from reflections on the different ways we write in our lives to the most ridiculous title we could conjure for our research. This was a productive way to get our writing engine revving and a fun opportunity to share writing with each other. Each evening we ended the day by a reflective discussion on how the day went (with an 'accountability buddy' [e.g. 'Rule 5' in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).

@@

Hosting a pre-retreat meeting with attendees before departure facilitates a smoother and more productive retreat by addressing the primary objectives of introductions, finalizing logistics, and goal-setting. During the event, participants will get a chance to meet (if they haven’t already) and establish or reinforce a familiar group dynamic, an essential prerequisite for a retreat atmosphere of respect and accountability (see Rules 5 and 10). Fine-tuning the plan for retreat logistics and scheduling through group conversation can identify opportunities for clarification and improvement, and create a sense of shared ownership over day-to-day programming. Most importantly, a pre-retreat meeting provides a chance for participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input of academic supervisors) so that these non-writing activities don't become distractions during the initial days of the retreat.

@@

A key aspect of a writing retreat is to have fun and enjoy the whole experience---from writing to exploring the local area. While on a writing retreat, it is important to take opportunities to experience new activities you may not usually get to. A retreat also presents a unique chance to bond with your peers while balancing downtime activities and writing time for optimal productivity.

Before you embark on your writing retreat, we recommend that you do some research before arriving to see what is unique about the area to take advantage of what may be on offer. While based at our university’s field station, we had opportunities to aurora hunt and kiwi spot, explore the nearby Ulva island (a protected area), participate in bar trivia, and develop relationships with some of the locals. We also organised creative morning writing prompts (see Rule 3) and gave casual presentations (Rule 7) about various topics which were fun and engaging bonding activities.

More rural places often have scenic landscapes, hospitable communities and limited distractions. This could restrict activity choices but may produce more quality experiences where you can get to know the local businesses and people on a more personal level. A retreat is not only about writing but an opportunity to engage with peers and have fun.

\section*{Rule 10: Posterity -- Be accountable and acknowledge}

Gathering feedback from participants of the retreat and drafting a report are important actions following the writing retreat. Allowing time for the participants to reflect on and discuss to their peers the pitfalls and peaks of the retreat on the last day or two of the retreat and then having them fill out a anonymous survey immediately after returning is very beneficial. By summarizing these points into a concise report soon after attending the retreat, points are clear in the minds of the participants and key issues or takeaways are not forgotten. Comments of the schedule, notes on things that went well, and notes of things that might need improvements are valuable to cement ideas and details that will fade in time before another retreat can be planned. This will reinforce great points from the retreat but more importantly creates some accountability for posterity. Knowledge transfer from one retreat to the next is difficult in an ever-changing student body, but creating a paper trail of advice and resources within your own department provides blueprints and templates for new volunteers to organize their own retreats. 
%The post-retreat report can act as a manual for future students and helps to lower the barrier to organizing a productive event by being a reference to future leaders with answers to FAQs and examples of how to organize administrative tasks.

In addition, give acknowledgement to the writing retreat in your work and publications and circulate this knowledge to your peers, your department, and your colleagues. Having acknowledgements of the retreat (potentially with a uniform phrase) can even provide a searchable tool in online databases. Creating a repository where work that has benefited from the writing retreat can be achieved by filtering these acknowledgements and compiling publications together in the same place. Highlighting to your department the quantity and quality of work that results from a writing retreat is key to securing the longevity of writing retreats. A list which updates into the future as new work is published that stemmed from the writing retreat means that administrative personnel can become aware of the significant impact this opportunity has provided student researchers. Both the scientific community and the university administration often use publication numbers as their tangible metric to interpret how their financial decisions have made an impact. Evidence of the how many papers may come out of a writing retreat by acknowledging the usefulness of it will help to capitalizing on this trend. A scientifically-productive retreat may ease the justification process of future student organizers, or even incentivise greater financial support from administrative groups if the expenditure of writing retreats are seen as a net-positive for publications.

@@

Following these set of rules, our writing retreats were sucessful. Colectivelly, we indeed experienced benefits reported in the literature, such as feeling empowered writers through supporting each other in a respectful and non-judgemental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara}] and decreasing our sense of isolation [~\cite{eardley2021}]. However, most importantly, we got our writing reatreat goals achieved. Some of the references herein can help with the theoretical side of writing, but ultimatelly writing is practice [~\cite{peterson2018, grogan2021}]. By facilitating the planning and execution of writing retreats, we hope this piece can foster scholarly productivity. Happy writing!

@@

\paragraph*{Writing – original draft:}  Nicholas W. Daudt (lead); Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S. Dixon-Anderson; Jackson Beagley; Joseph Curtis; Meghan Duffy; Saskia Foreman; Leah M. Crowe.

>>>>

Writing is an integral part of academic work. No matter in what career stage you are at, your scholarly output will advance scientific knowledge and count in your career development, getting grants or landing a job. However, in multi-tasking admin, teaching, fieldwork, among other activities, we frequently put aside how important it is to take the proper time to wrap it all up in a written document (e.g. thesis, scientific paper). Scientific writing has its technical tips and tricks, but ultimately, it is a creative act. Yet, the lack of time to think critically, create knowledge and transfer it into words is a significant complaint in academia [~\cite{menzies}]. Therefore, we have to actively disengage from other tasks to enable us to properly engage with the act of writing [~\cite{murray2013}]. In this context, writing retreats have been shown as effective opportunities to unplug from daily work routines, allowing scientists to excel in their passion: producing science.

@@

Here, we build on these experiences of student-led writing retreats and lay out Ten Simple Rules for organising an effective retreat. Althuogh the authors are/were all PhD candidates at the time of the retreats, the rules herein should be helpful for research groups and academics more broadly. We cover steps to facilitate the planning and execution of pre- (Rules 1--4), during (Rules 5--9), and pos- (Rule 10) writing retreat actions, but not writing techniques per se. There are many resources on academic writing that we will not delve into but encourage their reading (e.g. [~\cite{turbek2016, hotaling2020, carter2020, cargill2021}]), including articles in this series [~\cite{weinberger2015, mensh2017, peterson2018}]. Otherwise, you may consider writing a boring contribution instead [~\cite{sand2007}].

@@

The schedule for each day was roughly between 9 am and 6 pm and was provided in advance of the retreat so participants knew what to expect for dedicated work time (see \nameref{S1_Appendix}). A different facilitator was designated for each day to lead discussions and keep the time for writing blocks (between 60 and 120 min) and breaks (between 30 and 60 min). We started each morning together by writing for 15 minutes using prompts that varied from reflections on the different ways we write in our lives to the most ridiculous title we could conjure for our research. This was a productive way to get our writing engine revving and a fun opportunity to share writing with each other. Each evening we ended the day by a reflective discussion on how the day went (with an 'accountability buddy' [see Rule 5 in ~\cite{peterson2018}]), a shared dinner, and presentations (see Rule 7).

@@

Hosting a pre-retreat meeting with attendees before departure facilitates a smoother and more productive retreat by addressing the primary objectives of introductions, finalizing logistics, and goal-setting. During the event, participants will get a chance to meet (if they haven't already) and establish or reinforce a familiar group dynamic, an essential prerequisite for a retreat atmosphere of respect and accountability (see Rules 5 and 10). Fine-tuning the plan for retreat logistics and scheduling through group conversation can identify opportunities for clarification and improvement, and create a sense of shared ownership over day-to-day programming. Most importantly, a pre-retreat meeting provides a chance for participants to identify or workshop specific writing goals with time to prepare relevant resources (literature, data analysis, input of academic supervisors) so that these non-writing activities do not become distractions during the initial days of the retreat.

@@

A key aspect of a writing retreat is to have fun and enjoy the whole experience. While on a writing retreat, you might as well experience new activities you may not usually get to. A retreat also presents a unique chance to bond with your peers while balancing downtime activities and writing time for optimal productivity [~\cite{eardley2021, lyubykh}]. More rural places often have scenic landscapes, hospitable communities and limited distractions. This could restrict activity choices but may produce more quality experiences where you can get to know the local businesses and people on a more personal level. A retreat is not only about writing but an opportunity to engage with peers and have fun.

Before you embark on your writing retreat, we recommend that you do some research before arriving to see what is unique about the area to take advantage of what may be on offer. While based at our university's field station, we had opportunities to aurora hunt and kiwi spot, explore the nearby Ulva Island (a protected area), participate in bar trivia, and develop relationships with some of the locals. We also organised creative morning writing prompts (see Rule 3) and gave casual presentations (Rule 7) about various topics which were fun and engaging bonding activities.

\section*{Rule 10: Posterity---be accountable and acknowledge}

Gathering feedback from participants of the retreat and drafting a report are important actions following the writing retreat. Allowing time for the participants to reflect on and discuss to their peers the pitfalls and peaks of the retreat on the last day or two of the retreat and then having them fill out a anonymous survey immediately after returning is very beneficial. By summarizing these points into a concise report soon after attending the retreat, points are clear in the minds of the participants and key issues or takeaways are not forgotten. Comments of the schedule, notes on things that went well, and notes of things that might need improvements are valuable to cement ideas and details that will fade in time before another retreat can be planned. This will reinforce great points from the retreat but more importantly creates some accountability for posterity. Knowledge transfer from one retreat to the next is difficult in an ever-changing student body, but creating a paper trail of advice and resources within your own department provides blueprints and templates for new volunteers to organise their own retreats.

%The post-retreat report can act as a manual for future students and helps to lower the barrier to organizing a productive event by being a reference to future leaders with answers to FAQs and examples of how to organize administrative tasks. -- not sure if we did this, so maybe better to leave it out?

In addition, give acknowledgement to the writing retreat in your work and publications and circulate this knowledge to your peers, your department, and your colleagues. \textcolor{red}{Having acknowledgements of the retreat (potentially with a uniform phrase) can even provide a searchable tool in online databases.} Creating a repository where work that has benefited from the writing retreat can be achieved by filtering these acknowledgements and compiling publications together in the same place. Highlighting to your department the quantity and quality of work that results from a writing retreat is key to securing the longevity of writing retreats. A list which updates into the future as new work is published that stemmed from the writing retreat means that administrative personnel can become aware of the significant impact this opportunity has provided student researchers. Both the scientific community and the university administration often use publication numbers as their metric to interpret how their financial decisions have made an impact. Evidence of the how many papers may come out of a writing retreat by acknowledging the usefulness of it will help to capitalizing on this trend. A scientifically-productive retreat may ease the justification process of future student organisers, or even incentivise greater financial support from administrative groups if the expenditure of writing retreats are seen as a net-positive for publications.

@@

Following these set of rules, our  were sucessful.

Colectivelly, we experienced the benefits of writing retreats reported in the literature, such as feeling empowered writers through supporting each other in a respectful and non-judgemental environment [~\cite{papen2018}], increasing our sense of belonging as PhD students [~\cite{omeara}] and decreasing our sense of isolation [~\cite{eardley2021}]. However, most importantly, we got our writing goals for those weeks achieved. Some of the references herein can help with the theoretical side of writing, but ultimatelly writing is practice [~\cite{peterson2018, grogan2021}]. By facilitating the planning and execution of writing retreats through this Ten Simple Rules, we hope to foster scholarly productivity and wellbeing in student cohorts. Happy writing!

@@

\paragraph*{Writing – original draft:}  Nicholas W. Daudt; Claudia Hird; Eleanor Kelly; Gretchen McCarthy; Ian S. Dixon-Anderson; Jackson Beagley; Joseph Curtis; Meghan Duffy; Saskia Foreman; Leah M. Crowe.
